---
title: "SA_Project"
author: "Davide"
date: "2025-10-01"
output:
  pdf_document:
    fig_width: 4
    fig_height: 2.5
  html_document:
    df_print: paged
  word_document: default
  html_notebook: default
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
# Opzioni globali per knitr
knitr::opts_chunk$set(echo = FALSE, results = 'show')
# knitr::opts_chunk$set(fig.width=3.5, fig.height=2)
knitr::opts_chunk$set(fig.align = 'center')

```


## Exploratory Data Analysis
```{r}
library(ggplot2)
library(survival)
library(knitr)
library(dplyr)
library(survival)
library(ggplot2)
library(tibble)
library(lubridate)
```



```{r}

# Skip the first 12 lines of metadata
data <- read.csv("DirtSlurper3100.csv",
               skip = 12,
               stringsAsFactors = FALSE,
               na.strings = "---")

# Inspect
head(data)



```

Convert Dates
```{r}
data$Registration.date <- dmy(data$Registration.date)
data$Failure.date <- dmy(data$Failure.date)

head(data)
```



1. Registration Date Analysis

```{r}


ggplot(data, aes(x = Registration.date)) +
  geom_histogram(binwidth = 30,  
                 fill = "black", color="hotpink") +
  labs(title = "Registrations Over Time",
       x = "Registration Date", y = "Count")


```
2. Total Usage Time Analysis

```{r}

ggplot(data, aes(x = Total.usage.time)) +
  geom_histogram(binwidth = 30,  
                 fill = "pink", color="purple") +
  labs(title = "Distribution of usage",
       x = "Total Usage Time", y = "Count")

```

```{r}

ggplot(data, aes(x = Registration.date, y = Total.usage.time)) +
  geom_point(  alpha=0.2, color="hotpink") +
  labs(title = "Registrations Over Time",
       x = "Registration Date", y = "Usage Time(hours)")


```

We can see that the usage time highly depends on  when the registration is which makes sense but also is just proof that the data is right-censored. It s not a causal relationship, just stands to show how the data is cut at a certain point.

3. Pets

```{r}
ggplot(data, aes(x = Pets, y = Total.usage.time)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Usage Time by Pets", y = "Usage Time (hours)")

```
Surprising that the overall useage time seems to be higher for households with pets

4. Sent for repair
```{r}
ggplot(data, aes(x = Sent.for.repair, y = Total.usage.time)) +
  geom_boxplot(fill = "red") +
  labs(title = "Usage Time by repair", y = "Usage Time (hours)")

```
So I could say that the repairs look successful since the repaired ones were used lonegr on average

5. Carpet score
```{r}
data$Carpet.score <- as.factor(data$Carpet.score)

ggplot(data, aes(x = Carpet.score, y = Total.usage.time)) +
  geom_boxplot(fill = "darkgreen") +
  labs(title = "Usage Time by carpet", y = "Usage Time (hours)")

```

Interesting that most have the same statistics besides score 8. Maybe there are fewer scores 8?

```{r}
table(data$Carpet.score)
carpet_counts <- as.data.frame(table(data$Carpet.score))
colnames(carpet_counts) <- c("Carpet.score", "Count")
carpet_counts


```

So indeed there are very little instances of carpet with score 8 as well as 7. Scores 6 and 1 are also not particularly common.


6. Other
```{r}
data$time.to.failure <- as.numeric(data$Failure.date - data$Registration.date)

ggplot(data, aes(x = time.to.failure)) +
  geom_histogram(binwidth = 100, fill = "red", color = "black") +
  labs(title = "Time to Failure Distribution", x = "Days until Failure")


```
The days until failure(I do not think my plot accounts for NA values so I guess the NA values are counted as 0?) seem to be following a more or less normal distribution.



Kaplan Meier curve
```{r}
# Event: 1 if failure date present, 0 if missing (censored)
data$EventCensor = ifelse(is.na(data$Failure.date), 0L, 1L)

# Ensure time is numeric
data$Total.usage.time = as.numeric(data$Total.usage.time)

# Drop rows with missing time or event
df_km = subset(data, !is.na(Total.usage.time) & !is.na(EventCensor))

# KM Curve
library(survival)
KM1 = survfit(Surv(time = Total.usage.time, event = EventCensor) ~ 1, data = df_km)
plot(KM1, xlab = "Usage time", ylab = "Survival probability", main = "Kaplan–Meier curve")


```

Weibull only numerical, not estimating parameters
(Maybe in the end we can use the shape parameter of the Weibull)

Comparison with distributions and KM
KM also density function

Greenwood formula to inference (maybe later to include)



